# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
Repository comes with data compressed in file **activity.zip**.  Inside the zip archive, file **activity.csv** contains the data. It is possible to read fetch data directly in **data** dataframe without unarchiving the zip file:

```{r}
data <- read.csv(unz("./activity.zip","activity.csv"))
```


## What is mean total number of steps taken per day?
Histogram of the total number of steps taken each day indicates the dataset includes days with "0" and/or NA entries.
```{r}
dailySteps <- tapply(data$steps,data$date,sum)
barplot(dailySteps, col="blue", main ="Daily steps taken", ylab="Daily steps", xlab="Day")
```

Mean and median calculations must therefore take into acount presence of NA's values.
```{r, results='hide'}
mean_result <- mean(dailySteps,na.rm=TRUE)
median_result <- median(dailySteps,na.rm=TRUE)
```

Above calcualations give output values:
- mean of total steps taken per day: `r as.character(round(mean_result))`
- median of total steps taken per day: `r as.character(round(median_result))`

## What is the average daily activity pattern?
The pattern of average daily steps indicates most days are in the range of 30-50 steps:
```{r}
averageDailySteps <- tapply(data$steps,data$date,mean)
plot(averageDailySteps, type="l", main ="Pattern of average daily steps", ylab="Average Daily steps", xlab="Day")
points(averageDailySteps, pch=19)
```

In order to identify which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps; dataframe are recasted as follows:
```{r,warning=FALSE, echo=FALSE}
library(reshape2)
data$variable <- "steps"
maxDailySteps <- dcast(data=data,date~variable, fun.aggregate=max, value.var="steps", na.rm=TRUE)
maxDailySteps <- merge(maxDailySteps,data)
```

From **interval_table** new dataframe, vector of intervals with the maximum number of steps per each day is extracted and stored in **max_interval**, together with number of occurences in **max_days**:
```{r}
max_days <- max(table(maxDailySteps$interval))
max_interval <- names(which.max(table(maxDailySteps$interval)))
```
Code above leads to this result:
- In `r max_days` (ie. "max_days") different days maximum number of daily steps occurs during interval `r max_interval` (ie. "max_interval")


Interval that in **hours:minutes** format is equivalent to 5 minutes interval starting from **`r paste0(floor(as.numeric(max_interval)/60),":",as.numeric(max_interval)-(floor(as.numeric(max_interval)/60)*60))`**



## Imputing missing values
Count of NA's values gives the following result:
```{r}
sum(is.na(data$steps))
```

Strategy chosen to impute the missing values is replacing NA's value with the mean value of all the same 5-minutes intervals across all the days. Strategy is based on the assumption that number of steps take during the day are subject to a certain degree of *seasonality* during the 24 hours.

In order to fill the NA's average values for each interval are calculated across al days:
```{r}
avg_intervals <- dcast(data=data,interval~variable, fun.aggregate=mean, value.var="steps", na.rm=TRUE)
```

New dataframe **simulatedData** is then generated and simultanoeusly prefilled with average values for each interval.
```{r}
simulatedData <- merge(data[,c("interval","date")],avg_intervals)
```

New dataframe is then re-arranged according to sorting of original dataframe and its **steps** values are overwritten with original steps values where available in original **data** dataframe:
```{r}
library(plyr)
simulatedData <- arrange(simulatedData,date,interval)
availableSteps <- !is.na(data$steps)
simulatedData$steps[availableSteps] <- data$steps[availableSteps]
```

Histogram of the total number of steps taken each day is now.
```{r}
simulatedDataDailySteps <- tapply(simulatedData$steps,simulatedData$date,sum)
par(mfrow=c(2,1))
barplot(simulatedDataDailySteps, col="wheat", main ="Daily steps taken (simulated data)", ylab="Daily steps", xlab="Day")
barplot(dailySteps, col="blue", main ="Daily steps taken (original data)", ylab="Daily steps", xlab="Day")
```

Mean and median calculations are now.
```{r, results='hide'}
simulatedData_mean_result <- mean(simulatedDataDailySteps)
simulatedData_median_result <- median(simulatedDataDailySteps)
```
Above calcualations give output values for **simualted data**:
- mean of total steps taken per day: `r as.character(round(simulatedData_mean_result))`
- median of total steps taken per day: `r as.character(round(simulatedData_median_result))`

Two above values for **original data** were:
- mean of total steps taken per day: `r as.character(round(mean_result))`
- median of total steps taken per day: `r as.character(round(median_result))`

## Are there differences in activity patterns between weekdays and weekends?
To answer this question, simulated data are enriched with two columns defining 
- the name of the day corresponding to date
- The type of day "weekend" or "weekday"

```{r}
library(lubridate)
simulatedData$day <- wday(simulatedData$date,label=TRUE,abbr=FALSE)
simulatedData$day_type <- ifelse(simulatedData$day == "Sunday" | simulatedData$day == "Saturday","weekend","weekday")
```

Average step values for intevals during a weekday and during weekend days are calculcated and plotted
```{r}
averageSimulatedData <- dcast(simulatedData, interval ~ day_type, fun.aggregate=mean, value.var="steps")
averageSimulatedData <- melt(averageSimulatedData, id="interval",variable.name="day_type",value.name="steps")
library(ggplot2)
qplot(interval,steps, data=averageSimulatedData, facets=day_type~., geom="line", xlab="interval", ylab="No. of steps")
```

Results are interesting as there is a clear difference for intervals of **morning**:

- from 500 to 750 (ie. from **`r paste0(floor(500/60),":",500-(floor(500/60)*60))`** - to **`r paste0(floor(750/60),":",750-(floor(750/60)*60))`**)

where there are more steps taken during weekdays, probably due to work. Whilst the opposite is true for intervals of **late afternoon**:
- from 1000 to 2000 (ie. from **`r paste0(floor(1000/60),":",1000-(floor(1000/60)*60))`** - to **`r paste0(floor(2000/60),":",2000-(floor(2000/60)*60))`**)


